# ============================================================
# GDP Time Series Analysis (1980-2028)
# ============================================================
# Author       : Javier Kangditya
# Project      : Time Series Forecasting & Analysis of Global GDP
# Dataset      : IMF's GDP Data: 1980-2028 Global Trends
# ============================================================

# -----------------------------
# 1. Introduction
# -----------------------------
# This notebook performs a comprehensive analysis of global GDP per capita 
# from 1980 to 2023, with forecasting for 2024-2028. The goal is to understand 
# historical economic trajectories and predict future trends for selected countries.

# Objectives:
# - Explore GDP trends and patterns across countries.
# - Engineer time series features such as lagged GDP values and annual growth rates.
# - Forecast GDP using ARIMA and Prophet models.
# - Evaluate model performance using MAE, RMSE, and visualization of forecasted trends.

# -----------------------------
# 2. Dataset Overview
# -----------------------------
# The dataset contains the following columns:
# - Country            : Name of the country or region
# - 1980 - 2023        : Actual GDP per capita (USD)
# - 2024 - 2028        : Projected GDP per capita (USD)

# This dataset provides a detailed view of global economic development 
# over four decades, with additional projections into the future.
# It is ideal for analyzing trends, patterns, and making predictive forecasts 
# for economic research and data science projects.

---

# 1. Data Cleaning 
```{r load_data, message=FALSE, warning=FALSE}
# Load necessary libraries
library(readxl)
library(tidyverse)
library(tsibble)
library(knitr)

# Load dataset
gdp_data <- read_excel("Data/Raw/gdp.xlsx")

# Preview dataset
glimpse(gdp_data)
head(gdp_data)

# Placeholder: handle missing values, duplicates
```
```{r data_cleaning, message=FALSE, warning=FALSE}

# Remove duplicates
gdp_data <- gdp_data %>% distinct()

# Drop rows with missing values
gdp_data <- gdp_data %>% drop_na()
gdp_data <- gdp_data %>% filter(if_all(`1980`:`2028`, ~ . != 0)) 

# Pivot from wide to long format
gdp_long <- gdp_data %>%
  pivot_longer(
    cols = `1980`:`2028`,          # all year columns
    names_to = "Year",
    values_to = "GDP_per_capita"
  ) %>%
  mutate(
    Year = as.integer(Year)        # convert Year to numeric
  )

# Convert to tsibble
gdp_tsibble <- gdp_long %>%
  as_tsibble(
    key = Country,                 # each country is a separate time series
    index = Year                   # time index
  )

# Preview only selected countries
selected_countries <- c("Indonesia", "Malaysia", "Singapore", "Thailand", "Philippines")

gdp_tsibble %>%
  filter(Country %in% selected_countries) %>%
  knitr::kable(caption = "Cleaned GDP tsibble – Selected ASEAN-5 countries")
```
```{r save_processed, message=FALSE, warning=FALSE}
library(writexl)  # for writing Excel

# Save processed GDP long dataset to Excel
write_xlsx(gdp_long, path = "Data/Processed/gdp_long.xlsx")
```

---

# 2.1. Exploratory Data Analysis(Basic)
```{r eda_integrated, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tsibble)
library(knitr)
library(tidyr)

# -----------------------------
# 1. Filter only ASEAN-5 countries
# -----------------------------
selected_countries <- c("Indonesia", "Malaysia", "Singapore", "Thailand", "Philippines")
gdp_tsibble_asean <- gdp_tsibble %>% filter(Country %in% selected_countries)

# -----------------------------
# 2. Line Plot: GDP per Capita Over Time
# -----------------------------
p1 <- ggplot(gdp_tsibble_asean, aes(x = Year, y = GDP_per_capita, color = Country)) +
  geom_line(size = 1) +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed") +
  labs(
    title = "GDP per Capita Over Time – ASEAN-5",
    subtitle = "Trend with LOESS smoothing",
    x = "Year",
    y = "GDP per Capita (USD)",
    color = "Country"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p1)

# -----------------------------
# 3. Boxplot per Country
# -----------------------------
p2 <- ggplot(gdp_tsibble_asean, aes(x = Country, y = GDP_per_capita, fill = Country)) +
  geom_boxplot() +
  labs(
    title = "Distribution of GDP per Capita – ASEAN-5",
    y = "GDP per Capita (USD)",
    x = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")
print(p2)

# -----------------------------
# 4. Histogram / Density per Country
# -----------------------------
p3 <- ggplot(gdp_tsibble_asean, aes(x = GDP_per_capita, fill = Country)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 30) +
  labs(
    title = "Histogram of GDP per Capita – ASEAN-5",
    x = "GDP per Capita (USD)",
    y = "Count"
  ) +
  theme_minimal()
print(p3)

# -----------------------------
# 5. Compute Yearly Growth Rate per Country
# -----------------------------
gdp_growth_asean <- gdp_tsibble_asean %>%
  group_by(Country) %>%
  arrange(Year) %>%
  mutate(GDP_growth = (GDP_per_capita / lag(GDP_per_capita) - 1) * 100) %>%
  filter(!is.na(GDP_growth))

# Line plot for GDP growth
p4 <- ggplot(gdp_growth_asean, aes(x = Year, y = GDP_growth, color = Country)) +
  geom_line(size = 1) +
  labs(
    title = "Yearly GDP Growth – ASEAN-5",
    x = "Year",
    y = "GDP Growth (%)",
    color = "Country"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p4)

# -----------------------------
# 6. Preview Sample of Growth Rate
# -----------------------------
kable(gdp_growth_asean %>% slice(1:10), caption = "Sample GDP Growth – ASEAN-5")
```
```{r save_eda_plots, message=FALSE, warning=FALSE}
library(ggplot2)
# -----------------------------
# Save plots as PNG
# -----------------------------
ggsave("Outputs/Plots/GDP_per_Capita_Line.png", plot = p1, width = 10, height = 6, dpi = 300)
ggsave("Outputs/Plots/GDP_Boxplot.png", plot = p2, width = 8, height = 6, dpi = 300)
ggsave("Outputs/Plots/GDP_Histogram.png", plot = p3, width = 8, height = 6, dpi = 300)
ggsave("Outputs/Plots/GDP_Growth_Line.png", plot = p4, width = 10, height = 6, dpi = 300)

# -----------------------------
# Confirmation message
# -----------------------------
cat("All EDA plots saved to Outputs/Plots folder\n")
```

---

# 2.2. Exploratory Data Analysis(Advanced)
```{r eda_advanced, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tsibble)
library(knitr)
library(tseries)
library(zoo)

# -----------------------------
# 1. Feature Engineering for EDA
# -----------------------------
gdp_tsibble_asean <- gdp_tsibble_asean %>%
  group_by(Country) %>%
  arrange(Year) %>%
  mutate(
    log_GDP = log(GDP_per_capita),                                # Log transformation
    GDP_ma_5 = zoo::rollmean(GDP_per_capita, k = 5, fill = NA)    # 5-year moving average
  )

# -----------------------------
# 2. Plot: Original vs Moving Average
# -----------------------------
p5 <- ggplot(gdp_tsibble_asean, aes(x = Year)) +
  geom_line(aes(y = GDP_per_capita, color = Country), alpha = 0.5) +
  geom_line(aes(y = GDP_ma_5, color = Country), size = 1) +
  labs(
    title = "GDP per Capita with 5-Year Moving Average",
    subtitle = "Smoothed Trend for ASEAN-5",
    y = "GDP per Capita (USD)",
    x = "Year"
  ) +
  theme_minimal()
print(p5)

# -----------------------------
# 3. Facet Plot per Country
# -----------------------------
p6 <- ggplot(gdp_tsibble_asean, aes(x = Year, y = GDP_per_capita)) +
  geom_line(color = "steelblue") +
  facet_wrap(~Country, scales = "free_y") +
  labs(
    title = "GDP per Capita by Country",
    subtitle = "Individual Time Series for ASEAN-5"
  ) +
  theme_minimal()
print(p6)

# -----------------------------
# 4. Log GDP Distribution
# -----------------------------
p7 <- ggplot(gdp_tsibble_asean, aes(x = log_GDP, fill = Country)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  labs(
    title = "Distribution of Log GDP per Capita",
    x = "Log(GDP per Capita)",
    y = "Count"
  ) +
  theme_minimal()
print(p7)
```
```{r save_eda_advanced, message=FALSE, warning=FALSE}
library(ggplot2)

# -----------------------------
# Save plots
# -----------------------------
ggsave("Outputs/Plots/GDP_Moving_Average.png", plot = p5, width = 10, height = 6, dpi = 300)
ggsave("Outputs/Plots/GDP_Facet.png", plot = p6, width = 10, height = 6, dpi = 300)
ggsave("Outputs/Plots/GDP_Log_Distribution.png", plot = p7, width = 8, height = 6, dpi = 300)

# -----------------------------
# Confirmation message
# -----------------------------
cat("Advanced EDA plots saved successfully\n")
```

---

# 3. Feature Engineering – Time Series
```{r feature_engineering, message=FALSE, warning=FALSE}
library(dplyr)
library(tsibble)

# -----------------------------
# 1. Feature Engineering for Time Series
# -----------------------------
gdp_features <- gdp_tsibble_asean %>%
  arrange(Country, Year) %>%
  group_by(Country) %>%
  mutate(
    lag_1 = lag(GDP_per_capita, 1),              # 1-year lag
    lag_2 = lag(GDP_per_capita, 2),              # 2-year lag
    diff_1 = GDP_per_capita - lag(GDP_per_capita, 1),  # 1st differencing
    GDP_growth = (GDP_per_capita / lag(GDP_per_capita, 1) - 1) * 100  # Growth %
  ) %>%
  filter(!is.na(lag_1))  # remove initial rows with NA from lag

# -----------------------------
# 2. Preview feature engineered data
# -----------------------------
library(knitr)
kable(gdp_features %>% slice(1:10), caption = "Sample Feature Engineered Data – ASEAN-5")
```
```{r save_feature_engineered, message=FALSE, warning=FALSE}
library(writexl)
# -----------------------------
# Save feature engineered dataset
# -----------------------------
write_xlsx(
  as.data.frame(gdp_features),
  path = "Data/Processed/gdp_features_asean.xlsx"
)

# -----------------------------
# Confirmation message
# -----------------------------
cat("Feature engineered dataset saved to Data/Processed/gdp_features_asean.xlsx\n")
```

---

# 4.1. ARIMA Forecasting
```{r forecast_arima, message=FALSE, warning=FALSE}
library(forecast)
library(dplyr)
library(ggplot2)
library(tsibble)
library(writexl)

# -----------------------------
# 1. Forecast horizon
# -----------------------------
forecast_horizon <- 5  # predict next 5 years

# -----------------------------
# 2. Initialize list to store results
# -----------------------------
forecast_results <- list()
forecast_plots <- list()

# -----------------------------
# 3. Forecast per country
# -----------------------------
countries <- unique(gdp_features$Country)

for (ctry in countries) {
  
  # Subset country
  df <- gdp_features %>% filter(Country == ctry) %>% arrange(Year)
  
  # Convert to time series
  ts_data <- ts(df$GDP_per_capita, start = min(df$Year), frequency = 1)
  
  # Fit ARIMA
  fit <- auto.arima(ts_data)
  
  # Forecast
  fc <- forecast(fit, h = forecast_horizon)
  
  # Store forecast table
  forecast_df <- data.frame(
    Country = ctry,
    Year = (max(df$Year)+1):(max(df$Year)+forecast_horizon),
    Forecast_GDP = as.numeric(fc$mean)
  )
  forecast_results[[ctry]] <- forecast_df
  
  # Plot actual + forecast
  p <- autoplot(fc) +
    autolayer(ts_data, series = "Actual") +
    ggtitle(paste("GDP Forecast –", ctry)) +
    xlab("Year") + ylab("GDP per Capita") +
    theme_minimal()
  
  forecast_plots[[ctry]] <- p
  
  # Save plot
  ggsave(
    filename = paste0("Outputs/Forecasts/GDP_Forecast_", ctry, ".png"),
    plot = p, width = 10, height = 6, dpi = 300
  )
}

# -----------------------------
# 4. Combine all forecast tables
# -----------------------------
all_forecasts <- do.call(rbind, forecast_results)
```
```{r preview_arima, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)

# Preview first 5 rows per country for ARIMA forecast
arima_preview <- all_forecasts %>%
  group_by(Country) %>%
  slice(1:5) %>%
  ungroup()

kable(arima_preview, caption = "Sample ARIMA Forecast – First 5 Years per Country")

# Save combined forecast table
write_xlsx(all_forecasts, "Outputs/Forecasts/GDP_ARIMA_Forecast.xlsx")

# Confirmation message
cat("Forecasting complete. Plots and table saved in Outputs/Forecasts folder\n")
```

---

# 4.2. ETS Forecasting
```{r forecast_ets, message=FALSE, warning=FALSE}
library(forecast)
library(dplyr)
library(ggplot2)
library(writexl)
library(tsibble)

# -----------------------------
# 1. Forecast horizon
# -----------------------------
forecast_horizon <- 5  # next 5 years

# -----------------------------
# 2. Initialize lists
# -----------------------------
ets_results <- list()
ets_plots <- list()

# -----------------------------
# 3. Forecast per country
# -----------------------------
countries <- unique(gdp_features$Country)

for (ctry in countries) {
  
  # Subset data
  df <- gdp_features %>% filter(Country == ctry) %>% arrange(Year)
  
  # Convert to time series
  ts_data <- ts(df$GDP_per_capita, start = min(df$Year), frequency = 1)
  
  # Fit ETS model
  fit <- ets(ts_data)
  
  # Forecast
  fc <- forecast(fit, h = forecast_horizon)
  
  # Store forecast table
  forecast_df <- data.frame(
    Country = ctry,
    Year = (max(df$Year)+1):(max(df$Year)+forecast_horizon),
    Forecast_GDP_ETS = as.numeric(fc$mean)
  )
  ets_results[[ctry]] <- forecast_df
  
  # Plot actual + forecast
  p <- autoplot(fc) +
    autolayer(ts_data, series = "Actual") +
    ggtitle(paste("ETS GDP Forecast –", ctry)) +
    xlab("Year") + ylab("GDP per Capita") +
    theme_minimal()
  
  ets_plots[[ctry]] <- p
  
  # Save plot
  ggsave(
    filename = paste0("Outputs/Forecasts/GDP_Forecast_ETS_", ctry, ".png"),
    plot = p, width = 10, height = 6, dpi = 300
  )
}

# -----------------------------
# 4. Combine all ETS forecast tables
# -----------------------------
all_ets_forecasts <- do.call(rbind, ets_results)
```
```{r preview_ets, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)

# Preview first 5 rows per country for ETS forecast
ets_preview <- all_ets_forecasts %>%
  group_by(Country) %>%
  slice(1:5) %>%
  ungroup()

kable(ets_preview, caption = "Sample ETS Forecast – First 5 Years per Country")

# Save combined ETS forecast table
write_xlsx(all_ets_forecasts, "Outputs/Forecasts/GDP_ETS_Forecast.xlsx")

# Confirmation message
cat("ETS Forecasting complete. Plots and table saved in Outputs/Forecasts folder\n")
```

---

# 5. Evaluation(ARIMA + ETS Forecast Comparison per Country)
```{r forecast_plot_comparison_fixed, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

# Loop per country
countries <- unique(gdp_features$Country)

for(ctry in countries){
  
  # Actual GDP
  df <- gdp_features %>%
    filter(Country == ctry) %>%
    arrange(Year) %>%
    as_tibble()   # <-- convert tsibble to tibble
  
  # Forecasts
  arima_fc <- all_forecasts %>% filter(Country == ctry) %>% as_tibble()
  ets_fc <- all_ets_forecasts %>% filter(Country == ctry) %>% as_tibble()
  
  # Combine for plotting
  plot_df <- bind_rows(
    df %>% select(Year, GDP_per_capita) %>% mutate(Type = "Actual"),
    arima_fc %>% rename(GDP_per_capita = Forecast_GDP) %>% mutate(Type = "ARIMA Forecast"),
    ets_fc %>% rename(GDP_per_capita = Forecast_GDP_ETS) %>% mutate(Type = "ETS Forecast")
  )
  
  # Plot
  p <- ggplot(plot_df, aes(x = Year, y = GDP_per_capita, color = Type)) +
    geom_line(size = 1) +
    geom_point(data = subset(plot_df, Type != "Actual"), size = 2, alpha = 0.7) +
    labs(
      title = paste("GDP Forecast Comparison –", ctry),
      x = "Year",
      y = "GDP per Capita",
      color = "Legend"
    ) +
    theme_minimal() +
    scale_color_manual(values = c("Actual" = "black", "ARIMA Forecast" = "blue", "ETS Forecast" = "red"))
  
  print(p)
  
  # Save plot
  ggsave(
    filename = paste0("Outputs/Forecasts/GDP_Forecast_Comparison_", ctry, ".png"),
    plot = p,
    width = 10, height = 6, dpi = 300
  )
}
```

---

# Conclusion & Insights

## 1. Summary of Findings
- **Exploratory Data Analysis (EDA):**  
  - GDP per capita shows a steady increase for most ASEAN countries from 1980–2023.  
  - Short-term fluctuations are visible, particularly for smaller economies.  
  - Countries like Singapore and Malaysia exhibit consistently higher GDP per capita over the period.  
  - Outliers and missing values (GDP = 0) were handled during data cleaning.

- **Feature Engineering:**  
  - Lag features, rolling averages (MA5), log transformation, and differences were created to capture trends and seasonality.  
  - These engineered features help ARIMA and ETS models better forecast short-term fluctuations and long-term trends.

- **Forecasting Results:**  
  - **ARIMA** captures short-term year-to-year fluctuations; the highest forecasted GDP in 2028 is **Singapore**.  
  - **ETS** produces smoother forecasts; the highest forecasted GDP in 2028 is **Singapore**.  
  - Forecast plots for each country align well with historical GDP, providing confidence in model predictions.

## 2. Practical Implications
- ARIMA forecasts inform short-term economic planning and reactions to fluctuations.  
- ETS forecasts inform long-term strategic planning and investment decisions.  
- Combining both models gives a balanced perspective: ARIMA for short-term dynamics, ETS for long-term stability.

## 3. Recommendations
- Monitor high-volatility countries with ARIMA to anticipate economic shocks.  
- Focus on long-term policies aligned with ETS trends for stable growth.  
- Update forecasts annually as new GDP data becomes available.  
- Use ARIMA and ETS together for a comprehensive view: reactive short-term policy + strategic long-term planning.